\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2022-09-08 sections/discussion.tex
Update on Overleaf.

  Co-occurrence associations in microbial communities help identify important interdependencies that may drive or reflect microbial community structure and organization.
  A myriad of tools and methods have been developed for different parts of the workflow for inference of co-occurrence networks from 16S rRNA data.
  Our analysis shows that networks generated using different combinations of tools and approaches can be significantly different from each other, highlighting the need for a clear assessment of the source of variability and for tools that provide the most robust and accurate results.
  Our newly developed integrated software for the inference of co-occurrence networks from 16S rRNA data, \ac{micone}, is a freely customizable and user-friendly pipeline that allows users to compare networks generated by multiple possible combinations of tools and parameters.
  Importantly, in addition to revisiting the test cases presented in this work, users will be able to explore the effect of various tool combinations on their own datasets of interest.
  The \ac{micone} pipeline is built in a modular fashion; its plug-and-play architecture enables users to add new tools and steps, either using existing packages that were not examined in the present work or those developed in the future.
  The \ac{micone} Python package provides functions and methods to perform detailed analysis of the count matrices and the co-occurrence networks.
  The inferred networks are exported to a custom JSON format (see Supplementary) by default, but can also be exported to Cytoscape~\cite{shannonCytoscapeSoftwareEnvironment2003}, GML~\cite{himsoltGMLPortableGraph2010}, and many other popular formats via the \ac{micone} Python package.

  While several tools/workflows such as \ac{qiime2}~\cite{bolyenReproducibleInteractiveScalable2019} and NetCoMi~\cite{peschelNetCoMiNetworkConstruction2020} can be used to generate co-occurrence networks from 16S sequencing data, no single tool exist that integrates the complete process of inferring microbial interaction networks from 16S sequencing reads.
  \ac{micone} is unique as it offers this functionality packaged in a workflow that can be run locally, on the compute cluster or the cloud.

@@

  Through \ac{micone}, in addition to transparently revealing the dependence of co-occurrence networks on tool and parameter choices, we take advantage of our spectrum of computational options and the availability of mock and synthetic datasets, to suggest a default standard setting.
  Additionally, we develop a consensus approach, that can reliably generate networks that are robust across multiple tool/parameter choices.
  An important caveat related to these results is the fact that due to the lack of a universal standard for microbial interaction data, our conclusions are based on the specific datasets used in our analysis.
  While our datasets comprise several mock and synthetic datasets that cover a fairly diverse range of abundance distributions and network topologies, datasets that have drastically different distributions may require a re-assessment of the best settings through our pipeline.

  The extending discussion behind our choices and selection of default settings for the DC, TA and OP steps of the pipeline can be found in the supplementary discussion section (see Supplementary).
  The networks generated by different network inference methods show considerable differences in edge-density and connectivity.
  One reason for this is the underlying assumptions regarding sparsity, distribution and compositionality, that the algorithms make.
  The consensus network created by merging the networks inferred from the different network inference methods enables the creation of a network whose links have evidence based on multiple inference algorithms.
  We have developed two consensus algorithms to support this effort, the simple voting method and the scaled-sum method.
  Ultimately, we find that the scaled-sum method along with p-value merging (applied to the correlation-based methods) performs the best on synthetic datasets, therefore this is the default for the \ac{ni} step of the pipeline.
  The consensus algorithm not only leads to the creation of a network that has a higher number of true positives, it also decreases the variability in the networks due to choosing a different tool or parameter in any of the previous steps of the pipeline (Figure~\ref{fig:figure7}B).
  This provides the means to obtain a short list of associations, each of which, have a high probability of being present in the real interaction network
  This small set of associations could be further tested and validated experimentally to verify their plausibility.

@@

  Future work building upon our current results could enhance the network inference process in multiple ways.
  The current analyses make use of one fecal microbiome transplant dataset with healthy and autistic samples, three mock community datasets and several datasets generated by two synthetic interaction methods.
  Incorporating datasets from a broad spectrum of biomes with varying microbial distributions into \ac{micone} will likely increase the robustness and generalizability of the results from these analyses.

  The network analyses in this study are primarily at the Genus level, wherein the lowest resolution of a node is a Genus and if an entity cannot be resolved to the Genus level and next lowest taxonomic level is used (e.g. Family).
  As a consequence, two entities belonging to the same lineage where one entity is resolved to the Genus level and another is resolved to the Family level are treated as two different nodes in the network.
  Therefore, the development of a metric of overlap to compare nodes with shared lineages within and across networks could provide more biologically and phylogentically relevant comparisons.

  Further, benchmarking of co-occurrence networks could also be pursued through the utilization of literature-based interactions~\cite{lima-mendezDeterminantsCommunityStructure2015a} or biological benchmark interaction data~\cite{sungGlobalMetabolicInteraction2017a}.
  However, direct comparisons between co-occurrence networks and directly measured interactions are difficult to interpret and highly debated~\cite{hiranoDifficultyInferringMicrobial2019,gobernaCautionaryNotesUse2022}.
  Additionally, \ac{micone} could be extended to enable the processing of metagenomics sequencing data, facilitating the analysis of a much larger and diverse range of datasets and domains of life.

  Although in the current analysis, we only use default parameter values recommended by the tool creators, the \ac{micone} pipeline could be used in the future to explore any combinations of parameters and to optimize these values for improved network inference.
  Overall, it is important to keep in mind that there likely is no ``best method'' for the various steps of 16S data analysis, especially for the network inference step.
  Therefore, \ac{micone} is intended to help researchers to identify the methods and algorithms that are most suitable for their datasets in an easy-to-use and reproducible manner.

  We envision that \ac{micone}, and its underlying tools and databases, will be increasingly useful towards building large comparative analyses across studies.
  It allows for rapid, configurable and reproducible inference of microbial networks and further for the formulation of hypotheses about the role of these interactions on community composition and stability.
  These comparative analyses will require coupled network analysis and visualization tools (such as \ac{mind}~\cite{huResourceComparisonIntegration2022}), and will require a more systematic access to datasets, shared in accordance to FAIR standards~\cite{pachecoFAIRRepresentationsMicrobial2022}.

>>>>

  A myriad of tools and methods have been developed for different parts of the workflow for inference of co-occurrence networks from 16S rRNA data.
  Our analysis shows that networks generated using different combinations of tools and approaches can be significantly different from each other, highlighting the need for a clear assessment of the source of variability and for tools that provide the most robust and accurate results.
  Our newly developed integrated software for the inference of co-occurrence networks from 16S rRNA data, \ac{micone}, is a freely customizable and user-friendly pipeline that allows users to compare networks generated by multiple possible combinations of tools and parameters.
  Importantly, in addition to revisiting the test cases presented in this work, users will be able to explore the effect of various tool combinations on their own datasets of interest.
  The \ac{micone} pipeline is built in a modular fashion; its plug-and-play architecture enables users to add new tools and steps, either using existing packages that were not examined in the present work or those developed in the future.
  The \ac{micone} Python package provides functions and methods to perform a detailed analysis of the count matrices and the co-occurrence networks.
  The inferred networks are exported to a custom JSON format (see Supplementary) by default, but can also be exported to Cytoscape~\cite{shannonCytoscapeSoftwareEnvironment2003}, GML~\cite{himsoltGMLPortableGraph2010}, and many other popular formats via the \ac{micone} Python package.

  While several tools/workflows such as \ac{qiime2}~\cite{bolyenReproducibleInteractiveScalable2019} and NetCoMi~\cite{peschelNetCoMiNetworkConstruction2020} can be used to generate co-occurrence networks from 16S sequencing data, no single tool exist that integrates the complete process of inferring microbial interaction networks from 16S sequencing reads.
  \ac{micone} is unique as it offers this functionality packaged in a workflow that can be run locally, on the compute cluster, or in the cloud.

@@

  Through \ac{micone}, in addition to transparently revealing the dependence of co-occurrence networks on tool and parameter choices (see Discussion in Supplementary Text for details on the DC, TA and OP steps), we take advantage of our spectrum of computational options and the availability of mock and synthetic datasets, to suggest a default standard setting.
  Additionally, we develop a consensus approach, that can reliably generate networks that are robust across multiple tool/parameter choices.
  An important caveat related to these results is the fact that due to the lack of a universal standard for microbial interaction data, our conclusions are based on the specific datasets used in our analysis.
  While our analysis is based on several mock and synthetic datasets that cover a diverse range of abundance distributions and network topologies, datasets that have drastically different distributions may require a re-assessment of the best settings.

  The networks generated by different network inference methods show considerable differences in edge-density and connectivity, partially due to the underlying assumptions regarding sparsity, distribution and compositionality.
  To address this issue, we have developed two consensus algorithms (simple voting and scaled-sum method) that generate networks whose links have evidence based on multiple inference algorithms.

 We find that the scaled-sum method (along with p-value merging for correlation-based methods) performs the best on synthetic datasets, and is therefore chosen as default for the \ac{ni} step of the pipeline. Notably, the consensus network displays a higher number of true positives, while also decreasing the overall variability across the whole pipeline (Figure~\ref{fig:figure7}B). The concise list of robust associations revealed by the consensus algorithm represents a valuable set for experimental validation follow-up.

@@

  Future work building upon our current results could enhance the network inference process in multiple ways.
  The current analyses make use of one fecal microbiome transplant dataset with healthy and autistic samples, three mock community datasets, and several datasets generated by two synthetic interaction methods.
  Incorporating datasets from a broad spectrum of biomes with varying microbial distributions into \ac{micone} will likely increase the robustness and generalizability of the results from these analyses.

  The network analyses in this study are primarily at the Genus level, wherein the lowest resolution of a node is a Genus and if an entity cannot be resolved to the Genus level, the next lowest taxonomic level is used (e.g. Family).
  As a consequence, two entities belonging to the same lineage where one entity is resolved to the Genus level and another is resolved to the Family level are treated as two different nodes in the network.
  Therefore, the development of a metric of overlap to compare nodes with shared lineages within and across networks could provide more biologically and phylogenetically relevant comparisons.

  Although direct comparisons between co-occurrence networks and directly measured interactions are difficult to interpret and highly debated~\cite{hiranoDifficultyInferringMicrobial2019,gobernaCautionaryNotesUse2022}.
  Further, benchmarking of co-occurrence networks could also be pursued through the utilization of literature-based interactions~\cite{lima-mendezDeterminantsCommunityStructure2015a} or biological benchmark interaction data~\cite{sungGlobalMetabolicInteraction2017a}.
  Additionally, \ac{micone} could be extended to enable the processing of metagenomics sequencing data, facilitating the analysis of a much larger and diverse range of datasets and domains of life.

  Although in the current analysis, we only use default parameter values recommended by the tool creators, the \ac{micone} pipeline could be used in the future to explore any combinations of parameters and to optimize these values for improved network inference.
  Overall, it is important to keep in mind that there likely is no ``best method'' for the various steps of 16S data analysis.
  Therefore, \ac{micone} is intended to help researchers to identify the methods and algorithms that are most suitable for their datasets in an easy-to-use and reproducible manner.

  We envision that \ac{micone}, and its underlying tools and databases, will be increasingly useful for building large comparative analyses across studies.
  It allows for rapid, configurable, and reproducible inference of microbial networks and further for the formulation of hypotheses about the role of these interactions on community composition and stability.
  These comparative analyses will require coupled network analysis and visualization tools (such as \ac{mind}~\cite{huResourceComparisonIntegration2022}) and will require more systematic access to datasets, shared in accordance with FAIR standards~\cite{pachecoFAIRRepresentationsMicrobial2022}.

### 2022-09-08 sections/introduction.tex
Update on Overleaf.

  Microbial communities are ubiquitous and play an important role in marine and terrestrial environments, urban ecosystems, and human health \cite{Ghoul2016,Thompson2017}.
  These microbial communities, or microbiomes, often comprise several hundreds of different microbial strains interacting with each other and their environment, often through intricate metabolic and signaling relationships.
  Understanding how these interconnections shape community structure and functionalities is a fundamental challenge in microbial ecology, and has applications in the study of microbial ecosystems across different biomes.
  With the advancement in DNA sequencing technologies \cite{Narihiro2017} and data processing methods,  more information can be extracted from these microbial community samples than ever before.
  In particular, high-throughput sequencing, including metagenomic sequencing and sequencing of 16S rRNA gene amplicons (hereafter referred to as 16S data) of microbial communities, has the potential to help detect, identify and quantify a large portion of the constitutive microorganisms of a microbiome \cite{Jovel2016,Lloyd-Price2016}.
  These advances have led to large-scale data collection efforts involving environmental (\acl{emp}) \cite{Thompson2017}, marine (Tara Oceans Project) \cite{Zhang2015} and human-associated microbiota (Human Microbiome Project) \cite{HumanMicrobiomeProjectConsortium2012}.

 This wealth of information on the composition and functions of a community at different times and under different environmental conditions has the potential to help us understand how communities assemble and operate.
 A powerful tool for translating microbiome composition data into knowledge is the construction of possible inter-dependence networks across species.
 The importance of these networks of relationships is two-fold: first, such networks can serve as maps that help identify hubs of keystone species \cite{Menon2018,Rottjers2018}, and rewiring of interactions that occur as a consequence of environmental perturbations or underlying host conditions \cite{Gilbert2016}; second, networks of interdependencies can serve as a key bridge towards building mechanistic models of microbial communities, greatly enhancing our capacity to understand and control them.
 For example, multiple studies have shown the importance of specific microbial interactions in the healthy microbiome \cite{Lloyd-Price2016} and others have shown how changes in these interactions can lead to dysbiosis \cite{Wang2017,Gilbert2016,Belizario2015}.
 In the context of terrestrial bio-geochemistry, co-occurrence networks have been proposed as a valuable approach towards reconstructing the processes leading to microbiome assembly \cite{Fierer2017}, and understanding the response of microbial communities to environmental perturbations \cite{Jiao2019}.

 Direct high-throughput measurement of interactions, e.g. through co-culture micro-droplet experiments~\cite{Hsu2019,Jian2020}, observation of direct contact between cells~\cite{konovalovaCloseEncountersContactdependent2011}, or spatial visualization of natural communities \cite{shiHighlyMultiplexedSpatial2020,hartmannQuantitativeImageAnalysis2021,Wilbert2020} is possible, but it requires specific technological capabilities.
 However, sequencing data across multiple samples, which is more readily available, can be used for estimating co-occurrence relationships between taxa.
 While the relationship between directly measured interactions and statistically inferred co-occurrence is still poorly understood \cite{Zuniga2017}, a significant amount of effort has gone into estimating correlations from large microbiome sequence datasets.
 Co-occurrence networks have microbial taxa as nodes, and edges that represent the frequent co-occurrences (or negative correlations) across different datasets.

One of the most frequently used avenues for inferring co-occurrence networks is the parsing and analysis of 16S sequencing data \cite{Rottjers2018,Friedman2012}.
Numerous software tools and pipelines have been developed to analyze 16S sequencing data, often focused on addressing the many known limitations of this methodology, including resolution, sequencing depth, compositional nature, sequencing errors and copy number variations \cite{Bharti2019,Pollock2018}.
Popular methods for different phases of the analysis of 16S data include tools for: (i) quality checking and trimming the sequencing reads; (ii) denoising and clustering the trimmed reads \cite{Caporaso2010,Callahan2016}; (iii) assigning taxonomy to the denoised reads \cite{DeSantis2006,Quast2012}; (iv) processing and transforming the taxonomy count matrices \cite{Weiss2015}; and (v) inferring the co-occurrence network \cite{Cougoul2019,Kurtz2015}.
Different specific algorithms are often aggregated into popular online platforms (like MG-RAST~\cite{Keegan2016}, Qiita~\cite{qiita}) and software packages (such as QIIME2~\cite{bolyenReproducibleInteractiveScalable2019}) that provide pipelines for 16S data analysis.
The different methods and tools developed to solve issues arising in 16S analysis can lead to vastly different inferences of community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making it difficult to reliably compare networks across different publications and studies.
 This is partially due to the fact that existing platforms are typically focused on \ac{otu} generation and not on the effects of upstream statistical methods on the inferred co-occurrence networks.
 Furthermore, no organized framework currently exist to systematically analyze and compare existing components of the data analysis from amplicons to networks.
  More broadly, given the lack of comprehensive comparisons between directly observed microbial interactions (e.g. from co-culture experiments) and co-occurrence networks, there is no straightforward way to determine which set of tools or methods generate the most accurate networks.

 In this study, we present a standardized 16S data analysis pipeline called \ac{micone} that produces robust and reproducible co-occurrence networks from community 16S sequence data, and allow users to interactively explore how the network would change upon using different alternative tools and parameters at each step.
 Our pipeline is coupled to an online integrative tool for the organization, visualization and analysis of inter-microbial networks called \ac{mind}~\cite{huResourceComparisonIntegration2022}, that is available at \href{http://microbialnet.org/}{http://microbialnet.org/}.
  In addition to making this tool freely available, we completed a systematic comparative analysis to determine which steps of the pipeline have the largest influence on the final network, and what choice seems to have the best agreement with the tested mock and synthetic datasets.
  We believe that these steps will ensure better reproducibility and easier comparison of co-occurrence networks across datasets.
  We expect that our tool will also be useful for benchmarking future alternative methods, and for ensuring a transparent evaluation of the possible biases introduced by the use of specific tools.

>>>>

  Microbial communities are ubiquitous and play an important role in marine and terrestrial environments, urban ecosystems, and human health \cite{Ghoul2016,Thompson2017} [MORE REFS].
  These microbial communities, or microbiomes, often comprise several hundreds of different microbial strains interacting with each other and their environment, often through intricate metabolic and signaling relationships [REFS].
  Understanding how these interconnections shape community structure and functionalities is a fundamental challenge in microbial ecology, and has applications in the study of microbial ecosystems across different biomes.
  With the advancement in DNA sequencing technologies \cite{Narihiro2017} and data processing methods,  more information can be extracted from these microbial community samples than ever before.
  In particular, high-throughput sequencing, including metagenomic sequencing and sequencing of 16S rRNA gene amplicons (hereafter referred to as 16S data) of microbial communities, has the potential to help detect, identify and quantify a large portion of the constitutive microorganisms of a microbiome \cite{Jovel2016,Lloyd-Price2016}.
  These advances have led to large-scale data collection efforts involving terrestrial (\acl{emp}) \cite{Thompson2017}, marine (Tara Oceans Project) \cite{Zhang2015} and human-associated microbiota (Human Microbiome Project) \cite{HumanMicrobiomeProjectConsortium2012}. [3 REFS for each]

 This wealth of information has the potential to help us understand how communities assemble and operate. In particular, a powerful tool for translating microbiome composition data into knowledge is the construction of association (or co-occurrence, or correlation) networks, in which  microbial taxa are represented by nodes, and frequent co-occurrences (or negative correlations) across datasets are encoded in edges between nodes.
 While the relationship between directly measured interactions [REFS] and statistically inferred co-occurrence is still poorly understood \cite{Zuniga2017}, a significant amount of effort has gone into estimating correlations from large microbiome sequence datasets.

 The importance of these networks is two-fold: first, they can serve as maps that help identify hubs of keystone species \cite{Menon2018,Rottjers2018}, and the community response to environmental perturbations or underlying host conditions \cite{Gilbert2016}; second, they can serve as a bridge towards building mechanistic models of microbial communities, greatly enhancing our capacity to understand and control them.
 For example, multiple studies have shown the importance of specific microbial  associations in the healthy microbiome \cite{Lloyd-Price2016} and the rise of dysbiosis \cite{Wang2017,Gilbert2016,Belizario2015}.
 In the context of terrestrial bio-geochemistry, co-occurrence networks were shown to help understand microbiome assembly \cite{Fierer2017}, and  the response of microbial communities to environmental perturbations \cite{Jiao2019}.

One of the most frequently used avenues for inferring co-occurrence networks is the parsing and analysis of 16S sequencing data \cite{Rottjers2018,Friedman2012}.
Numerous software tools and pipelines have been developed to analyze 16S sequencing data, with strong emphasis on the known limitations of this methodology, including resolution, sequencing depth, compositional nature, sequencing errors and copy number variations \cite{Bharti2019,Pollock2018}.
Popular methods for different phases of the analysis of 16S data include tools for: (i) quality checking and trimming the sequencing reads; (ii) denoising and clustering the trimmed reads \cite{Caporaso2010,Callahan2016}; (iii) assigning taxonomy to the denoised reads \cite{DeSantis2006,Quast2012}; (iv) processing and transforming the taxonomy count matrices \cite{Weiss2015}; and (v) inferring the co-occurrence network \cite{Cougoul2019,Kurtz2015}.
Different specific algorithms are often aggregated into popular online platforms (like MG-RAST~\cite{Keegan2016}, Qiita~\cite{qiita}) and software packages (such as QIIME2~\cite{bolyenReproducibleInteractiveScalable2019}).
The different methods and tools developed to solve issues arising in 16S analysis can lead to vastly different inferences of community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making it difficult to reliably compare networks across different publications and studies.
 This is partially due to the fact that existing platforms are typically focused on \ac{otu} or \ac{esv} generation and not on the effects of upstream statistical methods on the inferred co-occurrence networks.
 Furthermore, no organized framework currently exist to systematically analyze and compare existing components of the data analysis from amplicons to networks.

 In this study, we present a standardized 16S data analysis pipeline called \ac{micone} that produces robust and reproducible co-occurrence networks from community 16S sequence data, and allow users to interactively explore how the network would change upon using different alternative tools and parameters at each step.
 Our pipeline is coupled to an online integrative tool for the organization, visualization and analysis of inter-microbial networks called \ac{mind}~\cite{huResourceComparisonIntegration2022}, that is available at \href{http://microbialnet.org/}{http://microbialnet.org/}.
Through a systematic comparative analysis we determined which steps of the \ac{micone} pipeline have the largest influence on the final network, and what choice seems to have the best agreement with the tested mock and synthetic datasets.
These steps will ensure better reproducibility and easier comparison of co-occurrence networks across datasets.
We expect that our tool will also be useful for benchmarking future alternative methods, and for ensuring a transparent evaluation of the possible biases introduced by the use of specific tools.
